%!TEX root = main.tex
\section{Models}

In this section, we introduce two one-dimensional phase-field fracture models for a brittle thin film bonded to substrates with different mechanical properties. These models provide a framework to investigate the stability and evolution of crack patterns under external loading.

\subsection{Thin Film on Stiff Substrate}
\label{sec:thin_film_stiff}

The first model describes a brittle thin film deposited on a stiff, non-deformable substrate. This model assumes the substrate is rigid, meaning it does not undergo any deformations regardless of the loading conditions. The energy functional for this system is constructed by considering the elastic energy of the thin film and the energy associated with the phase-field variable that describes the damage state of the film.
\newcommand{\Estiff}{\Psi}
\newcommand{\Ecompl}{\widetilde\Psi}
The total energy functional $\Estiff(\alpha, u)$ in the nondimensional form is given by:

nondimensional quantities:

\subsection{Thin Film on Compliant Substrate}
\label{sec:thin_film_compliant}
The second model involves a brittle thin film on a compliant, elastic substrate that can undergo non-uniform deformations. Unlike the rigid substrate in the first model, this substrate deforms under loading, which affects the overall energy landscape of the system.

The total energy functional $\Ecompl(\alpha, u, v)$ for this system is given by:







-------


In the subsequent section, we analyze the linear and nonlinear stability of these models. We will explore the equilibrium configurations and identify the conditions under which the system transitions between different equilibrium states. This analysis will involve both analytical and numerical methods, with a focus on understanding the implications of stability and bifurcation in these phase-field fracture models.